---
title: "457finalproject"
output: html_document
---

Group members: Sung Jegal, Minseo Kim, Paul Hwang

```{r}
library(dplyr)

G <- read.csv("Ghana_data.csv", header = TRUE, sep = ',', stringsAsFactors = FALSE)
C <- read.csv("Daily Prices_ICCO.csv", header = TRUE, sep = ',', stringsAsFactors = FALSE)
# plot(C, type= "l", ylab = "ICCO.daily.price..US..tonne.", xlab ="Date")
C$Date <- as.Date(C$Date, format = "%d/%m/%Y")  # Assuming day/month/year format

# First rename the column to something easier (optional but cleaner)
colnames(C)[colnames(C) == "ICCO.daily.price..US..tonne."] <- "Price"

# Remove commas and convert to numeric
C$Price <- gsub(",", "", C$Price)   # Remove commas
C$Price <- as.numeric(C$Price)      # Convert to numeric

# Verify it's numeric now
summary(C$Price)

C$Date <- as.Date(C$Date, format="%d/%m/%Y")

# Confirm structure
str(C$Date)
summary(C$Date)
```
```{r}
# Rename column DATE to Date
G1 <- G %>%
  rename(Date = DATE)


# Convert both Date columns to Date class
C$Date <- as.Date(C$Date, format="%d/%m/%Y")   # C: day/month/year
G1$Date <- as.Date(G1$Date, format="%Y-%m-%d")   # G: year-month-day

# Confirm they are both Date objects
class(C$Date)  # Should return "Date"
class(G1$Date)  # Should return "Date"

# If you need them *displayed* in a consistent format
C$Date_formatted <- format(C$Date, "%Y-%m-%d")
G1$Date_formatted <- format(G1$Date, "%Y-%m-%d")

# View the output
head(C$Date_formatted)
head(G1$Date_formatted)

```
```{r}
# Define the common date range
start_date <- as.Date("1994-10-03")
end_date   <- as.Date("2024-11-28")

# Filter Dataset C to this range
C_filtered <- subset(C, Date >= start_date & Date <= end_date)

# Filter Dataset G to this range
G_filtered <- subset(G1,  Date >= start_date & Date <= end_date)

# Check the ranges after filtering
range(C_filtered$Date)
range(G_filtered$Date)


# Merge datasets by Date
merged_df <- left_join(C_filtered, G_filtered, by="Date")

# Check the merged data
head(merged_df)
tail(merged_df)
nrow(merged_df)
```

```{r}
# install.packages("tidyr")

library(tidyr)
library(dplyr)

merged_df_filled <- merged_df %>% 
  fill(PRCP, TMAX, TMIN, .direction = "down")

merged_df_filled_1 <- na.omit(merged_df_filled)
head(merged_df_filled_1)
```
```{r}
# Assuming your dataset is merged_df_filled_1
daily_avg_df <- merged_df_filled_1 %>%
  group_by(Date) %>%
  summarise(
    avg_Price = mean(Price, na.rm = TRUE),
    avg_PRCP = mean(PRCP, na.rm = TRUE),
    avg_TAVG = mean(TAVG, na.rm = TRUE),
    avg_TMAX = mean(TMAX, na.rm = TRUE),
    avg_TMIN = mean(TMIN, na.rm = TRUE)
  ) %>%
  ungroup()
```

#Exploratory data analysis
```{r}
library(ggplot2)
library(forecast)

# Trajectory plot of average price over time
ggplot(daily_avg_df, aes(x = Date, y = avg_Price)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Trajectory Plot of Average Price over Time",
    x = "Date",
    y = "Average Price"
  ) +
  theme_minimal()

# plot acf and pacf
# Convert avg_Price to a time series object (optional but good practice)
price_ts <- ts(daily_avg_df$avg_Price)

# Plot ACF
acf(price_ts, main = "ACF of Average Price")

# Plot PACF
pacf(price_ts, main = "PACF of Average Price")

# qq-plot
qqnorm(daily_avg_df$avg_Price,
       main = "Q-Q Plot of Average Price")
qqline(daily_avg_df$avg_Price, col = "red")

# Differencing
# First difference
d_price <- diff(daily_avg_df$avg_Price)

# Plot first difference
plot(d_price, type = "l", main = "First Difference of Average Price", ylab = "Difference", xlab = "Time")

# ACF of first difference
acf(d_price, main = "ACF of First Difference of Average Price")

# PACF of first difference
pacf(d_price, main = "PACF of First Difference of Average Price")

# Second difference
dd_price <- diff(d_price)

# Plot second difference
plot(dd_price, type = "l", main = "Second Difference of Average Price", ylab = "2nd Difference", xlab = "Time")
acf(dd_price, main = "ACF of Second Difference of Average Price")
pacf(dd_price, main = "PACF of Second Difference of Average Price")
```

```{r}
# Load libraries
library(forecast)
library(tseries)
library(readr)
library(lubridate)
library(ggplot2)
library(dplyr)
library(astsa)

# Assume daily_avg_df is your cleaned dataset with the following columns:
# avg_Price, avg_PRCP, avg_TMAX, avg_TMIN
# Log-transform the average price
log_price <- log(daily_avg_df$avg_Price)

# Create exogenous variable matrix from the relevant columns
xreg_matrix <- as.matrix(daily_avg_df %>% select(avg_PRCP, avg_TAVG))

# Optional: check stationarity with ADF test if needed
#adf.test(log_price)

# Plot ACF and PACF of the log series
acf2(log_price, main="ACF/PACF of Log Price")

# Use auto.arima with exogenous variables to select a good model
#auto_model <- auto.arima(log_price)
#print(auto_model)

# Alternatively, manually fit an ARIMAX model; here we use order = (0,1,3) as an example.
model_arimax_log1 <- arima(log_price, order = c(1, 1, 0), xreg = xreg_matrix)
model_arimax_log2 <- arima(log_price, order = c(1, 1, 1), xreg = xreg_matrix)
model_arimax_log3 <- arima(log_price, order = c(2, 1, 1), xreg = xreg_matrix)

```
Residual Analysis ARIMAX (1,1,0)
```{r}

# Plot residuals
resid_model1 <- resid(model_arimax_log1)
plot(resid_model1, main = "Residuals of ARIMAX(1,1,0) Model", ylab = "Residuals")


#ACF of residuals
acf2(resid_model1, main = "ACF/PACF of Residuals of ARIMAX(1,1,0)")

# Extract coefficients
estimates <- model_arimax_log1$coef

# Extract standard errors (from the diagonal of the variance-covariance matrix)
std_errors <- sqrt(diag(model_arimax_log1$var.coef))

# Compute z-scores
z_scores <- estimates / std_errors

# Compute two-tailed p-values
p_values <- 2 * (1 - pnorm(abs(z_scores)))

# Combine results into a data frame
pval_table <- data.frame(
  Estimate = estimates,
  Std_Error = std_errors,
  Z_score = z_scores,
  P_value = p_values
)

# View the table
print(pval_table)

# Standardize the residuals, and analyse the standardized residuals
standardized_residuals1 <- resid(model_arimax_log1)/sqrt(model_arimax_log1$sigma2)
qqnorm(standardized_residuals1,
       main = "QQ plot of Standardized Residuals of ARIMAX(1,1,0)")
qqline(standardized_residuals1)

# Compute and plot Ljung-Box test p-values for lags 1 to 20
max_log <- 20
p_value <- numeric(max_log)
for (lag in 1:max_log) {
  p_value[lag] <- Box.test(standardized_residuals1, lag = lag, type = "Ljung-Box")$p.value
}

plot(1:max_log, p_value, type = "b",
     xlab = "Lag", ylab = "p-value", 
     main = "Ljung-Box Test p-values of Standardized Residuals of ARIMAX(1,1,0)",
     ylim = c(0, 1))
abline(h = 0.05, col = "blue", lty = 2)
```

Prediction ARIMAX (1,1,0)
```{r}
# Compute the in-sample fitted values (one-step-ahead forecasts)
fitted_values1 <- fitted(model_arimax_log1)
actual_values <- log_price
# Plot the actual log series in black
plot(log_price,
     type = "l",
     col = "black",
     main = "Log Price vs. One-step Ahead Forecasts of ARIMAX(1,1,0)",
     ylab = "Log of Series")

# Overlay the fitted (one-step-ahead) forecasts in red
lines(fitted_values1,
      col = "red",
      lty = 2)

# Legend 
legend("topleft",
       legend = c("Actual log_price", "One-step Ahead Fitted"),
       col = c("black", "red"),
       lty = c(1, 2))

# Calculate residuals
residuals <- actual_values - fitted_values1

# Compute RMSE
RMSE <- sqrt(mean(residuals^2))
print(paste("RMSE:", RMSE))

# Compute pseudo-R^2:
# SST is the total sum of squares: sum((actual - mean(actual))^2)
SST <- sum((actual_values - mean(actual_values))^2)
# SSE is the sum of squared errors: sum(residuals^2)
SSE <- sum(residuals^2)
R2 <- 1 - (SSE / SST)
print(paste("Pseudo R^2:", R2))


```
Residual Analysis ARIMAX (1,1,1)
```{r}

# Plot residuals
resid_model2 <- resid(model_arimax_log2)
plot(resid_model2, main = "Residuals of ARIMAX (1,1,1)", ylab = "Residuals")


#ACF of residuals
acf2(resid_model2, main = "ACF/PACF of Residuals ARIMAX (1,1,1)")


# Display AIC for the fitted ARIMAX model
#aic_value <- AIC(model_arimax_log2)
#print(paste("AIC:", aic_value))

# Extract coefficients
estimates <- model_arimax_log2$coef

# Extract standard errors (from the diagonal of the variance-covariance matrix)
std_errors <- sqrt(diag(model_arimax_log2$var.coef))

# Compute z-scores
z_scores <- estimates / std_errors

# Compute two-tailed p-values
p_values <- 2 * (1 - pnorm(abs(z_scores)))

# Combine results into a data frame
pval_table <- data.frame(
  Estimate = estimates,
  Std_Error = std_errors,
  Z_score = z_scores,
  P_value = p_values
)

# View the table
print(pval_table)

# Standardize the residuals, and analyse the standardized residuals
standardized_residuals2 <- resid(model_arimax_log2)/sqrt(model_arimax_log2$sigma2)
qqnorm(standardized_residuals2,
       main = "QQ plot of Standardized Residuals of ARIMAX(1,1,1)")
qqline(standardized_residuals2)

# Compute and plot Ljung-Box test p-values for lags 1 to 20
max_log <- 20
p_value <- numeric(max_log)
for (lag in 1:max_log) {
  p_value[lag] <- Box.test(standardized_residuals2, lag = lag, type = "Ljung-Box")$p.value
}

plot(1:max_log, p_value, type = "b",
     xlab = "Lag", ylab = "p-value", 
     main = "Ljung-Box Test p-values of Standardized Residuals of ARIMAX (1,1,1)",
     ylim = c(0, 1))
abline(h = 0.05, col = "blue", lty = 2)
```

Prediction ARIMAX (1,1,1)
```{r}
# Compute the in-sample fitted values (one-step-ahead forecasts)
fitted_values2 <- fitted(model_arimax_log2)
# Plot the actual log series in black
plot(log_price,
     type = "l",
     col = "black",
     main = "Log Price vs. One-step Ahead Forecasts ARIMAX (1,1,1)",
     ylab = "Log of Series")

# Overlay the fitted (one-step-ahead) forecasts in red
lines(fitted_values2,
      col = "red",
      lty = 2)

# Legend 
legend("topleft",
       legend = c("Actual log_price", "One-step Ahead Fitted"),
       col = c("black", "red"),
       lty = c(1, 2))

# Calculate residuals
residuals <- actual_values - fitted_values2

# Compute RMSE
RMSE <- sqrt(mean(residuals^2))
print(paste("RMSE:", RMSE))

# Compute pseudo-R^2:
# SST is the total sum of squares: sum((actual - mean(actual))^2)
SST <- sum((actual_values - mean(actual_values))^2)
# SSE is the sum of squared errors: sum(residuals^2)
SSE <- sum(residuals^2)
R2 <- 1 - (SSE / SST)
print(paste("Pseudo R^2:", R2))

```

Residual Analysis ARIMAX (2,1,1)
```{r}

# Plot residuals
resid_model3 <- resid(model_arimax_log3)
plot(resid_model3, main = "Residuals of ARIMAX Model", ylab = "Residuals")


#ACF of residuals
acf2(resid_model3, main = "ACF/PACF of Residuals ARIMA (2,1,1)")


# Display AIC for the fitted ARIMAX model
#aic_value <- AIC(model_arimax_log2)
#print(paste("AIC:", aic_value))

# Extract coefficients
estimates <- model_arimax_log3$coef

# Extract standard errors (from the diagonal of the variance-covariance matrix)
std_errors <- sqrt(diag(model_arimax_log3$var.coef))

# Compute z-scores
z_scores <- estimates / std_errors

# Compute two-tailed p-values
p_values <- 2 * (1 - pnorm(abs(z_scores)))

# Combine results into a data frame
pval_table <- data.frame(
  Estimate = estimates,
  Std_Error = std_errors,
  Z_score = z_scores,
  P_value = p_values
)

# View the table
print(pval_table)

# Standardize the residuals, and analyse the standardized residuals
standardized_residuals3 <- resid(model_arimax_log3)/sqrt(model_arimax_log3$sigma2)
qqnorm(standardized_residuals3,
       main = "QQ plot of Standardized Residuals of ARIMAX(1,1,0)")
qqline(standardized_residuals3)

# Compute and plot Ljung-Box test p-values for lags 1 to 20
max_log <- 20
p_value <- numeric(max_log)
for (lag in 1:max_log) {
  p_value[lag] <- Box.test(standardized_residuals3, lag = lag, type = "Ljung-Box")$p.value
}

plot(1:max_log, p_value, type = "b",
     xlab = "Lag", ylab = "p-value", 
     main = "Ljung-Box Test p-values of Standardized Residuals of ARIMAX (2,1,1)",
     ylim = c(0, 1))
abline(h = 0.05, col = "blue", lty = 2)
```

Prediction ARIMAX (2,1,1)
```{r}
# Compute the in-sample fitted values (one-step-ahead forecasts)
fitted_values3 <- fitted(model_arimax_log3)
actual_values <- log_price
# Plot the actual log series in black
plot(log_price,
     type = "l",
     col = "black",
     main = "Log Price vs. One-step Ahead Forecasts ARIMAX (2,1,1)",
     ylab = "Log of Series")

# Overlay the fitted (one-step-ahead) forecasts in red
lines(fitted_values3,
      col = "red",
      lty = 2)

# Legend 
legend("topleft",
       legend = c("Actual log_price", "One-step Ahead Fitted"),
       col = c("black", "red"),
       lty = c(1, 2))

# Calculate residuals
residuals <- actual_values - fitted_values3

# Compute RMSE
RMSE <- sqrt(mean(residuals^2))
print(paste("RMSE:", RMSE))

# Compute pseudo-R^2:
# SST is the total sum of squares: sum((actual - mean(actual))^2)
SST <- sum((actual_values - mean(actual_values))^2)
# SSE is the sum of squared errors: sum(residuals^2)
SSE <- sum(residuals^2)
R2 <- 1 - (SSE / SST)
print(paste("Pseudo R^2:", R2))

```


Compare AIC of the candidate models: we pick ARIMAX(2,1,1)
```{r}
# Display AIC for the fitted ARIMAX model
aic_value1 <- AIC(model_arimax_log1)
aic_value2 <- AIC(model_arimax_log2)
aic_value3 <- AIC(model_arimax_log3)

print(paste("AIC_110:", aic_value1))
print(paste("AIC_111:", aic_value2))
print(paste("AIC_211:", aic_value3))

```



```{r}
#install.packages("timeSeries")
#install.packages(c("timeDate", "timeSeries", "fBasics"))
#install.packages("fGarch")
#install.packages("rugarch")

library("astsa")
library("timeSeries")
library("fGarch")
library(tseries)
library(xts)
library(rugarch)
```

returns (from prof code)
```{r}
library(rugarch)
library(xts)    # optional if you're working with time-series objects


# Prepare data (growth rate / returns)
log_returns <- diff(log(daily_avg_df$avg_Price))

# Plot Returns
plot(log_returns, 
     type = "l", 
     main = "Log Returns of Cocoa Prices", 
     ylab = "Log Return")

# Step 3: ACF/PACF Returns
acf2(log_returns, 24, main = "ACF/PACF of Log Returns")
```


Fit MODELS
```{r}

X <- as.matrix(daily_avg_df %>% select(avg_PRCP, avg_TAVG))
# Check dimensions: they should match the length of your time series data
dim(X)


# Define your ARIMAX + GARCH specification
spec <- ugarchspec(
  variance.model = list(
    model = "sGARCH",        # standard GARCH model
    garchOrder = c(5, 1),
    external.regressors = NULL  # Optionally include regressors in variance
  ),
  mean.model = list(
    armaOrder = c(2, 1),       # ARMA order; adjust as needed
    include.mean = TRUE,
    external.regressors = X    # X is your regressor matrix for ARIMAX
  ),
  distribution.model = "norm"  # or "std" for Student's t-distribution, etc.
)

# Fit the model to your data (assume Y is your time series data)
GXmodel <- ugarchfit(spec, data = log_returns)

# To view the results:
# Get a detailed summary including coefficient significance:
# Extract the fitted model’s coefficient matrix
fit_coefs <- GXmodel@fit$matcoef

# Print it
fit_coefs
# Extract information criteria (like AIC) separately:
info <- infocriteria(GXmodel)
print(info)

summary(GXmodel)
```

Residual analysis of GXmodel (ARIMAX, GARCH)
```{r}
#ACF/PACF of GARCH
resdGXmodel = residuals(GXmodel)

plot(main = "ARMAX(2,1) + GARCH(5,1) residual",
     type = "l", 
     resdGXmodel)

acf2(resdGXmodel^2, main = "ACF/PACF of Residuals of ARMAX(2,1) + GARCH(5,1)")

# QQ-Plot of GARCH
GX_stnd_residuals <- residuals(GXmodel, standardize = TRUE)

qqnorm(GX_stnd_residuals,
       main = "QQ plot of Standardized Residuals of ARMAX(2,1) + GARCH(5,1)")
qqline(GX_stnd_residuals) # Deviation is noticeable but violation insignificant


# Compute and plot Ljung-Box test p-values for lags 1 to 20
max_log <- 20
p_value <- numeric(max_log)
for (lag in 1:max_log) {
  p_value[lag] <- Box.test(GX_stnd_residuals, lag = lag, type = "Ljung-Box")$p.value
}

plot(1:max_log, p_value, type = "b",
     xlab = "Lag", ylab = "p-value", 
     main = "Ljung-Box Test p-values of Standardized Residuals of ARMAX(2,1) + GARCH(5,1)", cex.main = 1,
     ylim = c(0, 1))
abline(h = 0.05, col = "blue", lty = 2)
```



Prediction for GXmodel
```{r}
forecast_obj <- ugarchforecast(GXmodel, n.ahead = 4)
#plot(forecast_obj)

# Extract actual log returns and fitted values from the model
fitted_valuesGX <- fitted(GXmodel)
actual_values <- daily_avg_df$avg_Price[-1]
last_four <- tail(actual_values, 4)

predictedGX <- exp(log(last_four) + forecast_obj@forecast$seriesFor)

print(predictedGX)

# Extract forecast sigma (volatility) on the log scale; this acts as the standard error for the forecasted log returns.
std_errors_log <- forecast_obj@forecast$sigmaFor

# Calculate standard errors on the original scale using the delta method.
# If Y = exp(X) then SE(Y) ≈ exp(X) * SE(X)
std_errors_orig <- predictedGX * std_errors_log

print(std_errors_orig)

log_vals <- log(actual_values) 
original_predictionGX = exp(fitted_valuesGX + log_vals)

# Extract actual price data (original scale)
n_data <- length(actual_values)

# PLOT: actual data in black, predicted in red
plot(
  1:n_data,               # x-axis indices
  actual_values,          # y values
  type = "l",
  col = "black",
  main = "Actual Prices vs. Four-step Ahead Fitted ARMAX(2,1) + GARCH(5,1) (Original Scale)", cex.main = 1,
  xlab = "Index",
  ylab = "Price",
  xlim = c(1, n_data + 5)
)

lines(
  1:n_data,                # same x indices
  original_predictionGX,        # fitted values on original scale
  col = "red",
  lty = 2
)

#lines(
#  n_data:n_data+3,                # same x indices
#  predictedGX,        # fitted values on original scale
#  col = "red",
#  lty = 2
#)

legend(
  "topleft",
  legend = c("Actual Price", "Four-step Ahead Fitted"),
  col = c("black", "red"),
  lty = c(1, 2),
  bty = "n"
)

# Compute RMSE
rmse <- sqrt(mean((log_returns - fitted_valuesGX)^2))
cat("RMSE:", rmse, "\n")

# Compute R² (pseudo R² for the mean equation)
ss_res <- sum((log_returns - fitted_valuesGX)^2)
ss_tot <- sum((log_returns - mean(log_returns))^2)
r_squared <- 1 - (ss_res / ss_tot)
cat("R²:", r_squared, "\n")



# Compute RMSE for original prices
rmse_prices <- sqrt(mean((actual_values - original_predictionGX)^2))
cat("RMSE for Original Prices of ARMAX(2,1) + GARCH(5,1):", rmse_prices, "\n")

# Compute R² for original prices:
ss_res_prices <- sum((actual_values - original_predictionGX)^2)
ss_tot_prices <- sum((actual_values - mean(actual_values))^2)
r2_prices <- 1 - ss_res_prices / ss_tot_prices
cat("R² for Original Prices of ARMAX(2,1) + GARCH(5,1):", r2_prices, "\n")


```





FIT MODEL GARCH
```{r}
#Fit GARCH model
#Gmodel1 <- garchFit(~arma(0, 0) + garch(1, 0), data = log_returns, trace = FALSE) #ARCH(1) 
Gmodel2 <- garchFit(~arma(2, 1) + garch(10, 1), data = log_returns, trace = FALSE) #GARCH(1,1)


#summary(Gmodel1)
summary(Gmodel2)
#summary(Gmodel3)
```

GARCH resdiual analysis
```{r}

#ACF/PACF of GARCH
resdGmodel2 = residuals(Gmodel2)

plot(main = "ARMA(2,1) + GARCH(10,1) residual",
     type = "l", 
     resdGmodel2)

acf2(resdGmodel2^2, main = "ACF/PACF of Residuals of ARMA(2,1) + GARCH(10,1)")

# QQ-Plot of GARCH
GARCH_stnd_residuals <- resdGmodel2/sqrt(Gmodel2@h.t)
qqnorm(GARCH_stnd_residuals,
       main = "QQ plot of Standardized Residuals of ARMA(2,1) + GARCH(10,1)", cex.main = 1)
qqline(GARCH_stnd_residuals) # Deviation is noticeable but violation insignificant


# Compute and plot Ljung-Box test p-values for lags 1 to 20
max_log <- 20
p_value <- numeric(max_log)
for (lag in 1:max_log) {
  p_value[lag] <- Box.test(GARCH_stnd_residuals, lag = lag, type = "Ljung-Box")$p.value
}

plot(1:max_log, p_value, type = "b",
     xlab = "Lag", ylab = "p-value", 
     main = "Ljung-Box Test p-values of Standardized Residuals of ARMA(2,1) + GARCH(10,1)", cex.main = 1,
     ylim = c(0, 1))
abline(h = 0.05, col = "blue", lty = 2)
```
GARCH prediction
```{r}
predict(Gmodel2, n.ahead = 4, plot = TRUE)


# Extract actual log returns and fitted values from the model
fitted_vals <- fitted(Gmodel2)
actual_valsG <- log_returns
log_vals <- log(actual_values) 
original_prediction = exp(fitted_vals + log_vals)


# Extract actual price data (original scale)
n_data <- length(actual_values)

# PLOT: actual data in black, predicted in red
plot(
  1:n_data,               # x-axis indices
  actual_values,          # y values
  type = "l",
  col = "black",
  main = "Actual Prices vs. Four-step Ahead Fitted ARMA(2,1) + GARCH(10,1) (Original Scale)", cex.main = 1,
  xlab = "Index",
  ylab = "Price",
  xlim = c(1, n_data + 5) 
)

lines(
  1:n_data,                # same x indices
  original_prediction,        # fitted values on original scale
  col = "red",
  lty = 2
)

legend(
  "topleft",
  legend = c("Actual Price", "Four-step Ahead Fitted"),
  col = c("black", "red"),
  lty = c(1, 2),
  bty = "n"
)

# Compute RMSE
rmse <- sqrt(mean((actual_valsG - fitted_vals)^2))
cat("RMSE:", rmse, "\n")

# Compute R² (pseudo R² for the mean equation)
ss_res <- sum((actual_valsG - fitted_vals)^2)
ss_tot <- sum((actual_valsG - mean(actual_valsG))^2)
r_squared <- 1 - (ss_res / ss_tot)
cat("R²:", r_squared, "\n")

#Reconvert Forecasted Returns to Original Series
# Get the last observed log price from your original data
last_log_price <- tail(log(daily_avg_df$avg_Price), 1)

# Compute cumulative sum of the forecasted returns to get the forecasted log prices
predicted_log_prices <- last_log_price + cumsum(fitted_vals)

# Convert forecasted log prices back to original price scale
predicted_prices <- exp(predicted_log_prices)


# Compute RMSE for original prices
rmse_prices <- sqrt(mean((actual_values - original_prediction)^2))
cat("RMSE for Original Prices:", rmse_prices, "\n")

# Compute R² for original prices:
ss_res_prices <- sum((actual_values - original_prediction)^2)
ss_tot_prices <- sum((actual_values - mean(actual_values))^2)
r2_prices <- 1 - ss_res_prices / ss_tot_prices
cat("R² for Original Prices:", r2_prices, "\n")

```




Fit MODELS ARCH(2)
```{r}
#Fit GARCH model
# ARCH(1) model specification

arch_spec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(2, 0)),
  mean.model = list(armaOrder = c(2, 1), include.mean = TRUE),
  distribution.model = "norm"
)

# Fit model
arch_fit <- ugarchfit(spec = arch_spec, data = log_returns)
show(arch_fit)

# 1. Forecast 5 steps ahead (from your fitted rugarch model)
forecast <- ugarchforecast(arch_fit, n.ahead = 5)

# 2. Plot forecasted returns (mean)
plot(forecast, which = 1)
title("Forecasted Log Returns (Mean Equation)", line = 3)

# 3. Plot forecasted volatility (sigma)
plot(forecast, which = 3, main = "Plot for the Forecasted volatility(sigma)")
title("Forecasted volatility (Sigma)", line = 3)
```

GARCH resdiual analysis
```{r}
return_fitted_vals <- fitted(arch_fit)
rmse <- sqrt(mean((log_returns - return_fitted_vals)^2))
cat("RMSE:", rmse, "\n")

actual_values <- daily_avg_df$avg_Price[-1]
# Compute RMSE for original prices
log_data = log(actual_values)
# Get starting index (skip NAs)
# original_fitted = exp(return_fitted_vals + log_data)

valid_idx <- which(!is.na(return_fitted_vals))
aligned_log_data <- log_data[(1 + length(log_data) - length(return_fitted_vals)):length(log_data)]

# Then compute original_fitted
original_fitted <- exp(return_fitted_vals + aligned_log_data)

rmse_prices <- sqrt(mean((actual_values - original_fitted)^2))
cat("RMSE for Original Prices:", rmse_prices, "\n")

# Compute R² for original prices
ss_res_prices <- sum((actual_values - return_fitted_vals)^2)
ss_tot_prices <- sum((actual_values - mean(actual_values))^2)
r2_prices <- 1 - ss_res_prices / ss_tot_prices
cat("R² for Original Prices:", r2_prices, "\n")

# Gmodel1 <- garchFit(~arma(0, 0) + garch(2 ,0), data = log_returns, trace = FALSE) #ARCH(1) 
Gmodel2 <- garchFit(~arma(2, 1) + garch(1, 1), data = log_returns, trace = FALSE) #GARCH(1,1)

#summary(Gmodel1)
summary(Gmodel2)
#summary(Gmodel3)
```

Prediction
```{r}
#Step 1: Forecast 100 steps ahead ===
forecast <- ugarchforecast(arch_fit, n.ahead = 10)

# Plot forecasted conditional mean (returns)
plot(forecast, which = 1)
title("Time series prediction (mean)", line = 2.5)

plot(forecast, which = 3)
title("Sigma prediction (volatility)", line = 2.5)

#Step 2: Residual diagnostics ===
resd_arch <- residuals(arch_fit, standardize = FALSE)

# Plot residuals
plot(resd_arch, 
     type = "l", 
     main = "ARMA(2,1) + GARCH(2,0) Residuals", 
     ylab = "Residuals")

# ACF/PACF of squared residuals
acf2(resd_arch^2, main = "ACF/PACF of Squared Residuals of ARMA(2,1) + GARCH(2,0)")

#Step 3: QQ-plot of standardized residuals ===
std_resid <- residuals(arch_fit, standardize = TRUE)

qqnorm(std_resid, 
       main = "QQ Plot of Standardized Residuals of ARMA(2,1) + GARCH(2,0)")
qqline(std_resid)

#Step 4: Ljung-Box p-values for lags 1 to 20 ===
max_lag <- 20
p_values <- numeric(max_lag)

for (lag in 1:max_lag) {
  p_values[lag] <- Box.test(std_resid, lag = lag, type = "Ljung-Box")$p.value
}

plot(1:max_lag, p_values, type = "b",
     xlab = "Lag", ylab = "p-value", 
     main = "Ljung-Box Test p-values of Standardized Residuals of ARMA(2,1) + GARCH(2,0)", cex.main = 1,
     ylim = c(0, 1))
abline(h = 0.05, col = "blue", lty = 2)
```

Fit model with regressor
```{r}
# Step 1: Prepare regressors (remove first row to match diff length)
xreg <- as.matrix(daily_avg_df[-1, c("avg_PRCP", "avg_TAVG")])

# Step 2: Specify the GARCH model with regressors
arch_spec_reg <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(15, 0)),
  mean.model = list(armaOrder = c(2, 1), include.mean = TRUE,
                    external.regressors = xreg),
  distribution.model = "norm"
)

# Step 3: Fit the model
arch_fit_reg <- ugarchfit(spec = arch_spec_reg, data = log_returns)
show(arch_fit_reg)

# Step 4: Forecast 5 steps ahead
forecast_reg <- ugarchforecast(arch_fit_reg, n.ahead = 5)

# Step 5: Plot mean forecast
plot(forecast_reg, which = 1)
title("Mean forecast Regressor", line = 3)
# Step 6: Plot volatility forecast
plot(forecast_reg, which = 3)
title("Volatility forecast (Regressor)", line = 3)

# Print
fit_coefs <- arch_fit_reg@fit$matcoef

fit_coefs

info <- infocriteria(arch_fit_reg)

print(info)
```

ARMA(2,1) + GARCH(2,0) Prediction

```{r}
# Extract actual log returns and fitted values from the model
fitted_valsA <- fitted(arch_fit)
actual_valsG <- log_returns
log_vals <- log(actual_values) 
original_predictionA = exp(fitted_valsA + log_vals)


# Extract actual price data (original scale)
n_data <- length(actual_values)

# PLOT: actual data in black, predicted in red
plot(
  1:n_data,               # x-axis indices
  actual_values,          # y values
  type = "l",
  col = "black",
  main = "Actual Prices vs. Four-step Ahead Fitted ARMA(2,1) + GARCH(2,0) (Original Scale)", cex.main = 1,
  xlab = "Index",
  ylab = "Price"
)

lines(
  1:n_data,                # same x indices
  original_predictionA,        # fitted values on original scale
  col = "red",
  lty = 2
)

legend(
  "topleft",
  legend = c("Actual Price", "Four-step Ahead Fitted"),
  col = c("black", "red"),
  lty = c(1, 2),
  bty = "n"
)
```


Residual analysis with regressor
```{r}
# Step 1: Extract fitted log returns
return_fitted_vals_reg <- fitted(arch_fit_reg)
return_fitted_vals_reg <- na.omit(return_fitted_vals_reg)  # Remove potential NA at start

# Step 2: Compute RMSE in log-return space
log_returns_aligned_reg <- tail(log_returns, length(return_fitted_vals_reg))
rmse_reg <- sqrt(mean((log_returns_aligned_reg - return_fitted_vals_reg)^2))
cat("RMSE_reg (Log-Returns):", rmse_reg, "\n")

# Step 3: Reconstruct predicted prices
log_price_data_reg <- log(daily_avg_df$avg_Price)
log_price_aligned_reg <- tail(log_price_data_reg, length(return_fitted_vals_reg))
actual_prices_reg <- exp(log_price_aligned_reg)
original_fitted_reg <- exp(return_fitted_vals_reg + log_price_aligned_reg)

# Step 4: Compute RMSE and R² in original price scale
rmse_prices_reg <- sqrt(mean((actual_prices_reg - original_fitted_reg)^2))
cat("RMSE_reg (Original Prices):", rmse_prices_reg, "\n")

ss_res_prices_reg <- sum((actual_prices_reg - original_fitted_reg)^2)
ss_tot_prices_reg <- sum((actual_prices_reg - mean(actual_prices_reg))^2)
r2_prices_reg <- 1 - ss_res_prices_reg / ss_tot_prices_reg
cat("R²_reg (Original Prices):", r2_prices_reg, "\n")
```

Prediction with regressor
```{r}
# Regressors for in-sample fit (must align with log_returns)
xreg_in_reg <- as.matrix(daily_avg_df[-1, c("avg_PRCP", "avg_TAVG")])
xreg_in_reg <- na.omit(xreg_in_reg)

# Regressors for forecast (last row repeated for 10-step forecast)
xreg_forecast <- matrix(rep(t(tail(xreg_in_reg, 1)), 10), ncol = 2, byrow = TRUE)

# === Step 2: Forecast 10 steps ahead ===
forecast_reg10 <- ugarchforecast(arch_fit_reg, 
                           n.ahead = 10, 
                           external.forecasts = list(mregfor = xreg_forecast))

# Plot forecasted conditional mean (returns)
plot(forecast_reg10, which = 1)
title("Forecasted conditional mean (returns) with regressor", line = 3)

# Plot forecasted conditional volatility (sigma)
plot(forecast_reg10, which = 3)
title("Forecasted conditional volatility (sigma) with regressor", line = 3)

# === Step 3: Residual diagnostics ===
resd_arch <- residuals(arch_fit_reg, standardize = FALSE)

plot(resd_arch, 
     type = "l", 
     main = "ARMAX(2,1) + GARCH(15,0) Residuals", 
     ylab = "Residuals")

acf2(resd_arch^2, main = "ACF/PACF of Squared Residuals of ARMAX(2,1) + GARCH(15,0)")

# === Step 4: QQ-plot of standardized residuals ===
std_resid_reg <- residuals(arch_fit_reg, standardize = TRUE)

qqnorm(std_resid_reg, 
       main = "QQ Plot of Standardized Residuals of ARMAX(2,1) + GARCH(15,0)")
qqline(std_resid_reg)

# === Step 5: Ljung-Box test for lags 1 to 20 ===
max_lag_reg <- 20
p_values_reg <- numeric(max_lag_reg)

for (lag in 1:max_lag) {
  p_values[lag] <- Box.test(std_resid_reg, lag = lag, type = "Ljung-Box")$p.value
}

plot(1:max_lag, p_values_reg, type = "b",
     xlab = "Lag", ylab = "p-value", 
     main = "Ljung-Box Test p-values of Standardized Residuals of ARMAX(2,1) + GARCH(15,0)", cex.main=1.0,
     ylim = c(0, 1))
abline(h = 0.05, col = "blue", lty = 2)
```

```{r}
# Extract the mean forecast
mean_forecast <- fitted(forecast)

# Extract forecasted sigma (volatility)
vol_forecast <- sigma(forecast)

# Suppose your original series was:
original_series <- fitted(arch_fit)  # or the original returns series you modeled

# Combine as a time series (xts or zoo is helpful here)
library(xts)

# Suppose your original_series is an xts object with a proper time index.
last_date <- index(original_series)[length(original_series)]
# Create a sequence of future dates; adjust the 'by' argument as needed (e.g., "days", "weeks")
future_dates <- seq(from = last_date + 1, by = "days", length.out = length(mean_forecast))
# Create the xts object for the forecast
forecast_xts <- xts(mean_forecast, order.by = future_dates)
# Combine original series with forecasted values
combinedAR <- c(original_series, forecast_xts)
combinedAR_og <- exp(combinedAR)

# Extract actual log returns and fitted values from the model
fitted_valsAR <- fitted(arch_fit_reg)
actual_valsG <- log_returns
log_vals <- log(actual_values) 
original_predictionAR = exp(fitted_valsAR + log_vals)

# Extract actual price data (original scale)
n_data <- length(actual_values)

# PLOT: actual data in black, predicted in red
plot(
  1:n_data,               # x-axis indices
  actual_values,          # y values
  type = "l",
  col = "black",
  main = "Actual Prices vs. Four-step Ahead Fitted ARMAX(2,1) + GARCH(15,0) (Original Scale)", cex.main = 1,
  xlab = "Index",
  ylab = "Price"
)

lines(
  1:length(combinedAR_og),                # same x indices
  combinedAR_og,        # fitted values on original scale
  col = "red",
  lty = 2
)

legend(
  "topleft",
  legend = c("Actual Price", "Four-step Ahead Fitted"),
  col = c("black", "red"),
  lty = c(1, 2),
  bty = "n"
)
```